---
title: "A tour of libraryStatistics"
author: "Anjali Silva"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{A tour of libraryStatistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility
```
       
## Introduction

`libraryStatistics` is an R package with a Shiny dashboard that permit to visualize and compare data from the annual survey of Association of Research Libraries (ARL; (www.arlstatistics.org/data/main)). ARL data describes the collections, staffing, expenditures, and service activities of the ARL member libraries in the United States and Canada. This R package is designed for the analysis and visualization of library statistics published from the annual survey. The Shiny application enables the generation of various statistical ratios for comparative analysis. Through the interactive interface of the application, users can dynamically visualize ratios by selecting ARL member libraries, years, and also by creating custom ratios, which enhances the usability of the published data. Library statistics data published from the annual survey may be downloaded from ARL Data Portal for any number of years and across any number of ARL member libraries. However, at one time, both the R package and Shiny application would only enable to perform analysis on 5 ARL member libraries and 5 years. **This document gives a tour of libraryStatistics (version 0.1.0)**. It was written in R Markdown, using the [knitr](https://cran.r-project.org/package=knitr) package for production. 

To download **libraryStatistics**, use the following commands:

``` r
require("devtools")
devtools::install_github("anjalisilva/libraryStatistics", build_vignettes = TRUE)
library("libraryStatistics")
```
To list all sample functions available in the package:
``` r
ls("package:libraryStatistics")
```

To list all sample datasets available in the package:
``` r
data(package = "libraryStatistics")
```

<br>


## Applications
The `libraryStatistics` R package contains 7 functions. Here we go through each:

1. __*shinyLibStats*__ opens the Shiny application with visual analytics and comparison of upto to 5 years of data from the annual survey of ARL uploaded by the user. Use the following commands to open Shiny application. 

```{r, }
library(libraryStatistics)
# libraryStatistics::shinyLibStats() # not run
``` 

The `libraryStatistics` Shiny application is a tool developed as part of the `libraryStatistics` R package. The Shiny application enables the generation of various statistical ratios for comparative analysis. Through the interactive interface of the application, users can dynamically visualize ratios by selecting ARL member libraries, years, and also by creating custom ratios, which enhances the usability of the published data.


2. __*visTotalLibraryExp*__ visualize total library expenditures in United States Dollars (USD) as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 6 on ARL survey  as the numerator.

To explore this function, we'll access the sample dataset of this package.

```{r, }
?ARLDataDownload
``` 

The `ARLDataDownload` is a demonstration dataset of size 18 rows and 80 columns, that shows the format required to be uploaded, as directly downloaded from ARL Data Portal, without any data preprocessing. The data file must be in a comma-separated value (.csv) format, where rows represent years and columns correspond to ARL indicators (variables). The first column should be labeled 'Year', followed by other indicators in any order, such as 'Institution Name', 'Institution type', etc., as directly downloaded from the ARL Data Portal. In total this example dataset contain data for 5 institutions, called Library A to Library E across three years from 2020, 2021 to 2022. The median value is also provided, as it is reflected in data downloaded from ARL Data Portal.

Using this dataset with __*libraryStatistics::visTotalLibraryExp()*__ function:
```{r, }
# outputvisTotalLibraryExp <- libraryStatistics::visTotalLibraryExp(dataARL = ARLDataDownload,
#                                      members = c("Library A", "Library B", "Library C"),
#                                      years = c(2020, 2021, 2022))
``` 

3. __*visTotalLibMaterialsExp*__ visualize total library materials expenditures in United States Dollars (USD) as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 7 on ARL survey as the numerator.


4. __*visProfStaffSalaries*__ visualize salaries of professional library staff in United States Dollars (USD), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 8a on ARL survey as the numerator.
5. __*visProfStaffCounts*__ visualize library professional staff counts, full-time equivalent (FTE), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 13a on ARL survey as the numerator.
6. __*visSupStaffCounts*__ visualize library support staff counts, full-time equivalent (FTE), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 13b on ARL survey as the numerator.
7. __*customRatioBuilder*__ build and visualize a custom ratio based on user
selected numerator and denominator from various statistics reported
in the annual survey of ARL.
8. __*indexTableGenerator*__ build a table containing Association of Research Libraries (ARL) Investment Index over years as reported in the
annual survey.


## Package References

- Silva, A. and K. Maidenberg (2024). libraryStatistics: An R Package with a Shiny Dashboard for Visualizing and Comparing Library Statistics Data from Association of Research Libraries. Unpublished.

<br>

## Other References

- R Core Team (2023). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/

- [Mian, A., & Gross, H. (2023). ARL Statistics 2022. Washington, DC: Association of Research Libraries.](https://publications.arl.org/ARL-Statistics-2022/) 

- [Association of Research Libraries. (2023). ARL Statistics 2023 Instructions.](https://www.arlstatistics.org/resources/stats_instructions)

----

```{r}
sessionInfo()
```
