---
title: "A tour of libraryStatistics"
author: "Anjali Silva"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{A tour of libraryStatistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility
```
       
## Introduction

`libraryStatistics` is an R package with a Shiny dashboard that permit to visualize and compare data from the annual survey of Association of Research Libraries (ARL; (www.arlstatistics.org/data/main)). ARL data describes the collections, staffing, expenditures, and service activities of the ARL member libraries in the United States and Canada. This R package is designed for the analysis and visualization of library statistics published from the annual survey. The Shiny application enables the generation of various statistical ratios for comparative analysis. Through the interactive interface of the application, users can dynamically visualize ratios by selecting ARL member libraries, years, and also by creating custom ratios, which enhances the usability of the published data. Library statistics data published from the annual survey may be downloaded from ARL Data Portal for any number of years and across any number of ARL member libraries. However, at one time, both the R package and Shiny application would only enable to perform analysis on 5 ARL member libraries and 5 years. **This document gives a tour of libraryStatistics (version 0.1.0)**. It was written in R Markdown, using the [knitr](https://cran.r-project.org/package=knitr) package for production. 

To download **libraryStatistics**, use the following commands:

``` r
require("devtools")
devtools::install_github("anjalisilva/libraryStatistics", build_vignettes = TRUE)
library("libraryStatistics")
```
To list all sample functions available in the package:
``` r
ls("package:libraryStatistics")
```

To list all sample datasets available in the package:
``` r
data(package = "libraryStatistics")
```

<br>


## Applications
The `libraryStatistics` R package contains 7 functions. Here we go through each:

1. __*shinyLibStats*__ function opens the Shiny application (app). The Shiny app enables the generation of various statistical ratios for comparative analysis. Through the interactive interface of the application, users can dynamically visualize ratios by selecting ARL member libraries, years, and also by creating custom ratios, which enhances the usability of the published data. Library statistics data published from the annual survey may be downloaded from ARL Data Portal for any number of years and across any number of ARL member libraries. However, at one time, both the R package and Shiny application would only enable to perform analysis on 5 ARL member libraries and 5 years. 

```{r, }
library(libraryStatistics)
# libraryStatistics::shinyLibStats() # not run
``` 

Steps:

i. Data Preparation: Begin by downloading the dataset from the ARL Data Portal (www.arlstatistics.org/data/main). Ensure that all variables are selected, with columns set to 'Variables' and the data sorted by 'Institution Name' (default options). Data should be downloaded in comma-separated value (.csv) format. Data may be downloaded for any number of years and across all member institutions available.

ii. Check Data: The downloaded dataset should have rows corresponding to years and columns to ARL indicators (variables). The first column must be 'Year', followed by other indicators in any order, such as 'Institution Name', 'Institution type', etc., as downloaded directly from the ARL Data Portal.

iii. Uploading Data and Parameter Selection: Upload the dataset (.csv format) to the Shiny application. After uploading the dataset, a list of choices for 'ARL Member Libraries' and 'Years' based on the uploaded dataset will appear. You may select up to 5 ARL member libraries and up to 5 years for analysis.

iv. Exploring Results: Navigate the tabs on the right side at the top of the application to explore the results. The left panel will remain static, allowing user to modify the selections for ARL member libraries or years, as needed. Changes to selections will automatically update the results displayed in the various tabs on the right.


2. __*visTotalLibraryExp*__ visualize total library expenditures in United States Dollars (USD) as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 6 on ARL survey  as the numerator of the ratio.

To explore this function, we'll access the sample dataset of this package.

```{r, }
?ARLDataDownload
``` 

The `ARLDataDownload` is a demonstration dataset of size 18 rows and 80 columns, that shows the format required to be uploaded, as directly downloaded from ARL Data Portal, without any data preprocessing. The format of the dataset is such that the rows represent years and columns correspond to ARL indicators (variables). The first column should be labeled 'Year', followed by other indicators in any order, such as 'Institution Name', 'Institution type', etc., as directly downloaded from the ARL Data Portal. In total this example dataset contain data for 5 institutions, called Library A to Library E across three years from 2020, 2021 to 2022. The median value is also provided, as it is reflected in data downloaded from ARL Data Portal.

Using this dataset with __*libraryStatistics::visTotalLibraryExp()*__ function:
```{r, }
# outputvisTotalLibraryExp <- libraryStatistics::visTotalLibraryExp(
#                                     dataARL = ARLDataDownload,
#                                     members = c("Library A", "Library B", "Library C"),
#                                     years = c(2020, 2021, 2022))
# outputvisTotalLibraryExp # To see bar plots
``` 

The output will  produce the following bar plots showing varying ratios specified below:

i. tlmeTopPerFaculty - A barplot showing ARL members with highest ratio of total library expenditures per teaching faculty, over user selected number of years.

ii. tlmeTopPerStudent - A barplot showing members with highest ratio of total library expenditures per student (full-time, FT, and part-time, PT), over user selected number of years.

iii. tlmeTopPerGradStudent - A barplot showing members with highest ratio of total library expenditures per graduate student (full-time, FT, and part-time, PT), over user selected number of years.

iv. tleTopPerUndergradStudent - A barplot showing members with highest ratio of total library expenditures per undergraduate student (full-time, FT, and part-time, PT), over user selected number of years.

v. tlmeTopPerDoctoral - A barplot showing members with highest ratio of total library expenditures per doctoral degree awarded, over user selected number of years.

vi. tlmePerFacultyUserSelected - A barplot showing ratio of total materials expenditures per teaching faculty for user selected ARL members, over user selected number of years.

vii. tlmePerStudentUserSelected - A barplot showing ratio of total materials expenditures per student (full-time, FT, and part-time, PT) for user selected ARL members, over user selected number of years.

viii. tlmePerGradStudentUserSelected - A barplot showing ratio of total materials expenditures per graduate student (full-time, FT, and part-time, PT) for user selected ARL members, over user selected number of years.

ix. tlePerUndergradStudentUserSelected - A barplot showing ratio of total materials expenditures per undergraduate student (full-time, FT, and part-time, PT) for user selected ARL members, over user selected number of years.

x. tlmeTopPerDoctoralUserSelected - A barplot showing ratio of total materials expenditures per per doctoral degree awarded for user selected ARL members, over user selected number of years.


3. __*visTotalLibMaterialsExp*__ visualize total library materials expenditures in United States Dollars (USD) as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 7 on ARL survey as the numerator of the ratio.
```{r, }
# outputvisTotalLibMaterialsExp <- libraryStatistics::visTotalLibMaterialsExp(
#                                     dataARL = ARLDataDownload,
#                                     members = c("Library A", "Library B", "Library C"),
#                                     years = c(2020, 2021, 2022))
# outputvisTotalLibMaterialsExp # To see bar plots

```

The output will produce following bar plots showing varying ratios specified below. The explanations are similar to ratios explained above. Also, see ?visTotalLibMaterialsExp.
i. tlmeTopPerFaculty
ii. tlmeTopPerStudent
iii. tlmeTopPerGradStudent
iv. tlmeTopPerUndergradStudent
v. tlmeTopPerDoctoral
vi. tlmePerFacultyUserSelected
vii. tlmePerStudentUserSelected
viii. tlmePerGradStudentUserSelected
ix. tlmePerUndergradStudentUserSelected
x. tlmePerDoctoralUserSelected

4. __*visProfStaffSalaries*__ visualize salaries of professional library staff in United States Dollars (USD), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 8a on ARL survey as the numerator of the ratio.

```{r, }
# outputvisProfStaffSalaries <- libraryStatistics::visProfStaffSalaries(
#                                     dataARL = ARLDataDownload,
#                                     members = c("Library A", "Library B", "Library C"),
#                                     years = c(2020, 2021, 2022))
# outputvisProfStaffSalaries # To see bar plots

```

The output will produce following bar plots showing varying ratios specified below. The explanations are similar to ratios explained above. Also, see ?visProfStaffSalaries.
i. proSalTopPerFaculty
ii. proSalTopPerStudent
iii. proSalTopPerGradStudent
iv. proSalTopPerUndergradStudent
v. proSalTopPerDoctoral
vi. proSalFacultyUserSelected
vii. proSalPerStudentUserSelected
viii. proSalPerGradStudentUserSelected
ix. proSalPerUndergradStudentUserSelected
x. proSalPerDoctoralUserSelected

5. __*visProfStaffCounts*__ visualize library professional staff counts, full-time equivalent (FTE), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 13a on ARL survey as the numerator.

```{r, }
# outputvisProfStaffCounts <- libraryStatistics::visProfStaffCounts(
#                                     dataARL = ARLDataDownload,
#                                     members = c("Library A", "Library B", "Library C"),
#                                     years = c(2020, 2021, 2022))
# outputvisvisProfStaffSalaries # To see bar plots

```

The output will  produce following bar plots showing varying ratios specified below. The explanations are similar to ratios explained above. Also, see ?visProfStaffCounts.
i. proFTETopPerFaculty
ii. proFTETopPerStudent
iii. proFTETopPerGradStudent
iv. proFTETopPerUndergradStudent
v. proFTETopPerDoctoral
vi. proFTEPerFacultyUserSelected
vii. proFTEPerStudentUserSelected
viii. proFTEPerGradStudentUserSelected
ix. proFTEPerUndergradStudentUserSelected
x. proFTEPerDoctoralUserSelected

6. __*visSupStaffCounts*__ visualize library support staff counts, full-time equivalent (FTE), as ratios in comparison to various statistics reported in the annual survey of ARL as bar plots. This is question 13b on ARL survey as the numerator.

```{r, }
# outputvisSupStaffCounts <- libraryStatistics::visSupStaffCounts(
#                                     dataARL = ARLDataDownload,
#                                     members = c("Library A", "Library B", "Library C"),
#                                     years = c(2020, 2021, 2022))
# outputvisSupStaffCounts # To see bar plots

```

The output will produce following bar plots showing varying ratios specified below. The explanations are similar to ratios explained above. Also, see ?visSupStaffCounts.
i. supFTETopPerFaculty
ii. supFTETopPerStudent
iii. supFTETopPerGradStudent
iv. supFTETopUndergradStudent
v. supFTETopPerDoctoral
vi. supPerFacultyUserSelected
vii. supPerStudentUserSelected
viii. supPerGradStudentUserSelected
ix. supPerUndergradStudentUserSelected
x. supPerDoctoralUserSelected

7. __*customRatioBuilder*__ build and visualize a custom ratio based on user
selected numerator and denominator from various statistics reported
in the annual survey of ARL.

```{r, }
# outputviscustomRatioBuilder <- libraryStatistics::customRatioBuilder(
#                 dataARL = ARLDataDownload,
#                 numerator = "Electronic books",
#                 denominator = "Total fulltime students",
#                 members = c("Institute A", "Institute B", "Institute C", "Institute D", "Institute E"),
#                 years = c(2020, 2021, 2022))

# outputviscustomRatioBuilder # To see bar plots

```

The output will produce following bar plots showing varying ratios specified below. Also, see ?customRatioBuilder.

i. customRatioTop - A barplot showing members with highest ratio of custom ratio based on user selected numerator and denominator, over user selected years.

ii. customRatioUser - A barplot showing ratio based on user selected numerator and denominator, for user selected ARL members, over user selected number of years.

8. __*indexTableGenerator*__ build a table containing Association of Research Libraries (ARL) Investment Index over years as reported in the
annual survey.

```{r, }
# outputIndexTableGenerator <- libraryStatistics::indexTableGenerator(
#                   dataARL = ARLDataDownload,
#                   members = c("Library A", "Library B", "Library C", "Library D", "Library E"),
#                   years = c(2020, 2021, 2022))


# outputIndexTableGenerator # To see bar plots

```

This returns a table of class knitr_kable showing Association of Research Libraries (ARL) Investment Index for all institutes in the uploaded data and for user selected years.

## Package References

- Silva, A. and K. Maidenberg (2024). libraryStatistics: An R Package with a Shiny Dashboard for Visualizing and Comparing Library Statistics Data from Association of Research Libraries. Unpublished.

<br>

## Other References

- R Core Team (2023). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/

- [Mian, A., & Gross, H. (2023). ARL Statistics 2022. Washington, DC: Association of Research Libraries.](https://publications.arl.org/ARL-Statistics-2022/) 

- [Association of Research Libraries. (2023). ARL Statistics 2023 Instructions.](https://www.arlstatistics.org/resources/stats_instructions)

----

```{r}
sessionInfo()
```
